Multi-level modelling with Openflexo/FML - A contribution to the \mpc

### 2021-05-21 2-Abstract.tex
OpenFlexo -> Openflexo

Model federation is a multi model \emph{management} approach based on the use of virtual models and loosely coupled links. The models in a federation remain autonomous and represented in their original technological spaces whereas virtual models and links serve as control components used to present different views to the users and  maintain synchronization. In this paper we tackle the \emph{process modeling challenge}, which consists in providing a solution to the problem of specifying and enacting processes. Solutions must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We present a solution based on model federation and discuss the advantages and limitations of using this approach for multi-level modeling. Our solution is fully implemented with the OpenFlexo framework and its internal Flexo Modeling Language (FML). It fulfills all the challenge requirements.

\noteSM{J'ai fait un premier essai d'abstract... un peu trop vendeur peut-etre?}
\noteJC{bof, pas de manière éhontée : il faut qu'un abstract soit un peu
  vendeur et les exigences sont bien remplies. La fin est très positive, mais
afactuelle selon moi.}

>>>>

Model federation is a multi model \emph{management} approach based on the use of virtual models and loosely coupled links. The models in a federation remain autonomous and represented in their original technological spaces whereas virtual models and links serve as control components used to present different views to the users and  maintain synchronization. In this paper we tackle the \emph{process modeling challenge}, which consists in providing a solution to the problem of specifying and enacting processes. Solutions must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We present a solution based on model federation and discuss the advantages and limitations of using this approach for multi-level modeling. Our solution is fully implemented with the Openflexo framework and its internal Flexo Modeling Language (FML). It fulfills all the challenge requirements.

\noteSM{J'ai fait un premier essai d'abstract... un peu trop vendeur peut-etre?}
\noteJC{bof, pas de manière éhontée : il faut qu'un abstract soit un peu
  vendeur et les exigences sont bien remplies. La fin est très positive, mais
afactuelle selon moi.}
\noteSylvain{Ce qui m'embête un peu avec cet abstract c'est qu'on met en avant la fédération de modèles alors que sur ce cas précis on en fait pas vraiment (sauf pour la partie outillage avec les éditeurs diagrammatiques). Concernant la fédération de modèle proprement dite, je crois qu'il faut qu'on montre qu'on a besoin d'une couche conceptuelle (qu'on relie à l'espace technique via les connecteurs techniques, les model slots et les rôles). C'est ce besoin qu'adresse le language FML. On montre ici plutôt que le language FML - utilisé comme language conceptuel - est assez intéressant, assez expressif pour capturer les exigences du challenge d'une part, et va jusqu'à permettre l'exécution des processus. Et cerise sur le gâteau: on peut construire un outillage graphique pour editer et exécuter ce use case.}
\noteSylvain{En revanche, on peut arguer que pouvoir au même niveau manipuler modèle et métamodèle nous simplifie grandement les choses pour adresser des problématiques multi-niveaux > un bout de la conclusion ?}.

### 2021-05-21 8-Discussion.tex

discussing choices  made (ex : héritage vs attr), pointing out potential compromises/deficiencies
jc, fabien ; 
pas d'explicitation de niveaux de « concrétude » : conséquences ? facile de tracer des liens entre les niveaux ; difficile d'avoir le niveau ; 
outil adaptable, niveau d'abstraction, on fournit une solution + l'outil pour exécuter le processus ; réutilisabilité des modèles ; 
\todo{on pourrait aussi répondre ici aux 3 questions posées explicitement (copiées dansl'intro pour le moment)}

difficulctés/facilités
points saillants :  outillage "gratuit" ;

Although our tool was not designed with a focus on multi-level modelling, we were able to propose a fully working solution.

Our approach is level-blind, therefore it is not easy to quickly retrieve the abstraction level of a concept.

>>>>

\noteJC{notes en vrac
  \begin{itemize}
    \item qui ? jc, fabien
    \item l'appel contraint plus ou monis la structure et le contenu de cette
      section
  \begin{itemize}
    \item discussing choices  made (ex : héritage vs attr), pointing out potential compromises/deficiencies ;
    \item \emph{Challenge respondents must discuss their multilevel
model solution with regard to the following aspects, each
of which should be treated in a specific sub-section of the
“Assessment” section of the article}
  \end{itemize}
    \item pas d'explicitation de niveaux de « concrétude » : conséquences ? facile de tracer des liens entre les niveaux ; difficile d'avoir le niveau ;
    \item outil adaptable
    \item niveau d'abstraction
    \item réutilisabilité des modèles ; 
    \item ajouter quelques métriques "gros grains"
    \item difficultés/facilités : contraintes faites à la main par rapport aux
      langages où on peut spécifier le niveau d'extension lorsque l'on trace un
      lien multi niveaux -> outil plus souple, mais moins de comportements
      automatisés
    \item + ouvert
    \item moins de support (faudrait implémenter ce qui manque)
    \item requiert de la méthode, ajouter des comportements, etc.
    \item point saillant :  outillage "gratuit" = on fournit une solution + l'outil pour exécuter le processus
\end{itemize}}

\todo[inline]{on pourrait aussi répondre ici aux 3 questions posées
explicitement (copiées dansl'intro pour le moment), a priori non, le chef
trouve que c'est chiant}

%
%%niveaux de ``concrétude'' -> level-blind solution
%
%Although our tool was not designed with a focus on multi-level modelling, we were able to propose a fully working solution. 
%
%Our approach is level-blind, therefore it is not easy to quickly retrieve the abstraction level of a concept.
%
%%métriques gros grains ?
%
%
%%choices, compromises/deficiences
%During the modeling phase, we had to choose 
%(for instance requirement xx 
%
%%+/-
%
%%points saillants

  %plan plus ou moins imposé par l'appel
\todo[inline]{mandatory discussion aspects (on verra plus tard si on fait des
sous-sections ou pas) ↓}

  \subsection{Basic modeling constructs}
  %Explain the basic modeling constructs used in the solution.

  \subsection{Levels}
  %Levels (or other model content organization schemes employed)

  %Explain the nature of “levels” in the model, how model elements are arranged
  %on these levels and which relationships (such as “instance-of”) may feature
  %between elements at different levels. The nature of levels should be
  %captured by explicitly stating the level segregation and the level cohesion
  %principles used [5]. Avoid vague language such as “higher level concepts are
  %more abstract” if the inter-level relationship is more specific. If the
  %inter-level relationship is deliberately allowed to be vague, state this
  %explicitly.

  The Openflexo approach is level-blind: ``levels'' have no specific
  nature and there are no numbered levels.

  \subsection{Number of levels}
  %Elaborate whether the submitted solution could have had more or fewer levels
  %and explain how any potentially existing degrees of freedom were resolved.

  Due to the fact our approach is level-blind, our solution could have had more
  or fewer levels depending on the variations of the use case. %As levels are
  %not specifically encoded,

  \subsection{Cross-level relashionships}

  %Discuss if and how associations and links can connect model elements at
  %different levels. State well-formedness constraints, if any apply.

  Thanks to the level-blindness nature of the Openflexo approach, cross-level
  relashionships are not an issue. Model elements of different levels can be
  linked each other in a transparent way. %If the user has defined
  %well-formedness constraints, then they are

  \subsection{Cross-level constraints}

%Discuss if and how constraints can span multiple levels, especially with
%regard to cross-level relationships.

  As for cross-level relationships, cross-level constraints do not have a
  specific nature. They are like any constraint a user can define on a model
  element. Therefore, if a constraint is mandatory when establishing a
  relationship between elements from different levels, the user has create it
  manually. This can be a limit of our approach: the flexibility of our tooling
  at the cost of a limited number fo automated behaviors.

  \subsection{Integrity mechanism}

  %Discuss how the integrity of level contents is preserved when changes to
  %level contents occur.

  In order to preserve the integrity of level contents when changes to level
  contents occur, a validation has to be triggered. The tool reports all
  inconsistencies that the user is free to fix.

  \subsection{Deep characterization}

  %Discuss if and how higher levels influence elements at lower levels with a
  %level distance of two or more. Such an influence may be desired to ensure
  %properties of lower level elements regardless of the design choices that
  %modelers make at intermediate levels, including future extensions to
  %intermediate levels.

  \noteJC{pas trop clair pour moi pour le moment}

  \subsection{Generality}

  %Discuss the generality of the solution. Is (part of) it applicable to other
  %domains? Does it embody invariant principles of the domain(s) it covers with
  %minimal redundancy?

  Due to the nature of model federation, the models are highly reusable. Our
  solution separates the domain-specific elements from general purpose ones,
  making easily possible to apply it to other domains. 
  \noteJC{à développer + ajouter avantage de l'outil gratuit ici ?}

  \subsection{Extensibility}

  %Elaborate how the solution would respond to further requirements, such as
  %further special tasks that must be taken care of by special actors. Identify
  %expected extension points in the solution, e.g., subtyping opportunities. If
  %level insertion is a possibility in your chosen approach, explain how it
  %would be performed.

  A strength of the approach of model federation we have adopted resides in its
  flexibility. One can extend a solution easily:

  \noteJC{comment étendre : dépend un peu des extensions, ne prendre que celle
  donnée en exemple ? ; level insertion possible : oui, vu que pas de notion de
level}

\todo[inline]{recommended discussion aspects ↓}
\subsection{formalisms and tools ?}

%Indicate whether there are formalisms to establish the semantics of the MLM
%technique and/or tools that support the present solution

\noteJC{on a du tooling, un peu moins de formalisme}

\subsection{model verification}

%Discuss model verification (e.g., consistency analyses) or other quality
%assessment mechanisms supported by the MLM technique employed.

### 2021-05-21 9-Relatedworks.tex
OpenFlexo -> Openflexo

\cite{multiecore2019} use MultiEcore \citep{multecore2016} to provide a solution to the challenge. MultEcore uses cascading and potency. They do not need the synthetic typing relations we introduce. However they need to do frequent model transformations to transform instance models into instantiable models. With respect to them, we use less models as we do not include a generic software engineering process model nor a generic assurance process model. Finally, they use multi-level couple model transformations for the definition and enforcement of constraints.

More similar to us, \cite{deeptelos2019} use DeepTelos\citep{deeptelos2016}, an extension of the Telos language \citep{telos1990} in order to solve the process challenge. DeepTelos is based on the use of the PowerType \citep{atkinson2001essence} pattern. As with OpenFlexo and FML, they do not use explicit (numbered) levels nor potency.

@@

\cite{dmla2019} Use the Dynamic Multi-Layer Algebra (DMLA)\citep{dmla2017} in order to solve the process challenge. DMLA is a level-blind modeling framework which is fully customizable (e.g., different types of instantiaton may be implemented). Their proposed solution separates the process challenge in two separate domains, namely, the tasks definition domain and the process definition domain. Wrappers are used in order to reuse tasks in processes. Their solution does not use inheritance, as it is not supported by the framework. It does not use explicitly separated models either.

>>>>

\cite{multiecore2019} used MultiEcore \citep{multecore2016} to provide a solution to the challenge. MultEcore uses cascading and potency. They do not need the synthetic typing relations we introduce. However they need to do frequent model transformations to transform instance models into instantiable models. With respect to them, we use less models as we do not include a generic software engineering process model nor a generic assurance process model. Finally, they use multi-level couple model transformations for the definition and enforcement of constraints.

More similar to us, \cite{deeptelos2019} used DeepTelos\citep{deeptelos2016}, an extension of the Telos language \citep{telos1990} in order to solve the process challenge. DeepTelos is based on the use of the PowerType \citep{atkinson2001essence} pattern. As with Openflexo and FML, they do not use explicit (numbered) levels nor potency.

@@

\cite{dmla2019} used the Dynamic Multi-Layer Algebra (DMLA)\citep{dmla2017} in order to solve the process challenge. DMLA is a level-blind modeling framework which is fully customizable (e.g., different types of instantiaton may be implemented). Their proposed solution separates the process challenge in two separate domains, namely, the tasks definition domain and the process definition domain. Wrappers are used in order to reuse tasks in processes. Their solution does not use inheritance, as it is not supported by the framework. It does not use explicitly separated models either.
